Structural rather than catalytic role for mitochondrial respiratory chain supercomplexes

Version 1

Abstract

Abstract

Mammalian mitochondrial respiratory chain (MRC) complexes are able to associate into quaternary structures named supercomplexes (SCs), which normally coexist with non-bound individual complexes. The functional significance of SCs has not been fully clarified and the debate has been centered on whether or not they confer catalytic advantages to the non-bound individual complexes. Mitochondrial respiratory chain organization does not seem to be conserved in all organisms. In fact, and differently from mammalian species, mitochondria from insect tissues are characterized by low amounts of SCs, despite the high metabolic demands and MRC activity shown by these mitochondria. Here, we show that attenuating the biogenesis of individual respiratory chain complexes was accompanied by increased formation of stable SCs, which are missing in Drosophila melanogaster in physiological conditions. This phenomenon was not accompanied by an increase in mitochondrial respiratory activity. Therefore, we conclude that SC formation is necessary to stabilize the complexes in suboptimal biogenetic conditions, but not for the enhancement of respiratory chain catalysis.

Introduction

Mitochondria are the organelles providing most of the cellular energy in form of adenosine triphosphate (ATP) in aerobic eukaryotes. The molecular machinery responsible for energy transformation is the oxidative phosphorylation (OXPHOS) system, which is canonically composed of five multiprotein complexes embedded in the inner mitochondrial membrane. OXPHOS consists of two tightly regulated processes: electron transport and ATP synthesis. Electron transport takes place between complexes I-IV and two mobile electron carriers (coenzyme Q and cytochrome c). During electron transport, complexes I, III and IV pump protons from the mitochondrial matrix to the intermembrane space, generating a proton gradient that provides the protonmotive force exploited by complex V to synthesize ATP. In mammalian mitochondria, mitochondrial respiratory chain (MRC) complexes I, III and IV can interact with each other forming supramolecular structures described generally by the term ‘supercomplexes’ 1,2. MRC supercomplexes (SCs) can have different stoichiometries and compositions, ranging from the binding of only two complexes, such as the I1III2 and III2IV1 SCs 3,4, to higher-order associations between complexes I, III and IV, with the SC of I1III2IV1 stoichiometry known as the ‘respirasome’ 1,5–9. Now that the association of individual MRC complexes into supramolecular structures is well-established, with structures of several SC species being resolved, the debate is centered on what the functional significance of these structures might be.

Several possible roles have been proposed for SCs. First, it was suggested that the association between complex I (CI) and the obligate dimer of complex III (CIII2) would allow substrate channeling, sequestering a dedicated coenzyme Q (CoQ) pool and allowing a more efficient electron transfer between the two complexes, while separating this electronic route from those of FADH2-linked dehydrogenases (e.g. complex II) to the CIII2 not bound to CI 1,10–15. This increased efficiency would in turn decrease electron leak and, as a consequence, produce less reactive oxygen species (ROS) than the individual free complexes 16,17. However, the available high-resolution respirasome structures show that the distance between the CoQ binding sites in CI and CIII2 are too far apart to allow substrate channeling 4,18. In addition, exogenously added CoQ was necessary to sustain CI activity in the purified I1III2 SC 3, arguing against a tightly bound and segregated CoQ pool as a functional component of the SC. A large body of work from the late 1960’s to the 1980’s, resulted in the ‘random collision model’ to explain electron transfer between the respiratory chain complexes, and in the evidence that CoQ is present as an undifferentiated pool19–23. More recently, additional proof of the non-compartmentalized electronic routes from CI to CIII2 and from complex II (CII) to CIII2, came from kinetic measurements in sub-mitochondrial particles. In these systems, MRC organization in SCs was conserved but b- and c-type cytochromes in CIII2 were equally accessible to CI-linked and CII-linked substrates 24, and CoQ reduced by CI in the respirasomes was able to reach and readily reduce external enzymes to the SCs 25. In addition, growing evidence supports the notion that different MRC organizations exist in vivo, where varying proportions of SC vs. free complexes do not result in separate and distinct CI-linked and CII-linked respiratory activities 26–29. This is in contrast with an idea that segregation into different types of SCs and in individual complexes to explains their functional interplay for the adaptation of OXPHOS activity 13. The physical proximity of CIII2 and CIV has also been suggested to promote faster electron transfer kinetics via cytochrome c 4,30,31, although this is a matter of debate as well 32,33.

The second main explanation to justify the existence of SCs is that they play a structural function, stabilizing the individual complexes 34,35 and/or serving as a platform for the efficient assembly of the complexes, with a special relevance for the biogenesis of mammalian CI 36–38.

Notably, the MRC structural organization, especially the stoichiometry, arrangements and stability of the SCs, may not be conserved in all eukaryotic species 39–42. This is the case even within mammals, as human cells and tissues barely contain free CI, which is rather contained in SC I1+III2 and the respirasome 29,37. In contrast, other mammalian mitochondria (from bovine, ovine, rat or mouse) contain larger amounts of CI in its free form, even if the majority is still mostly in the form of SCs 2,17,28, 43–45. The distribution of the MRC complexes between free complexes and SCs seems to differ even more in non-mammalian animal species. Several reptile species contain a very stable SC I+III2 that lacks CIV 28, and in Drosophila melanogaster practically all of CI is free, SCs being almost completely absent 46,47. However, comparative studies of MRC function in diverse animal species suggest that higher amounts and stability of the SCs do not correlate with increased respiratory activity/efficiency and/or reduced ROS production 28,47.

Here we show that SCs can be stably formed in D. melanogaster mitochondria upon mild perturbations of individual CIV, CIII2 and CI biogenesis. This finding enabled us to test whether increased SC formation translated into enhanced respiration proficiency. However, MRC performance of fruit fly mitochondria did not change regardless of the presence or absence of SCs. These observations have led us to conclude that: 1) the efficiency in the assembly of the individual complexes is likely to be the main determinant of SC formation and 2) these supramolecular complexes play a more relevant role in maintaining the stability and/or supporting the biogenesis of the MRC than in promoting catalysis.

Results

D. melanogaster MRC organization does rely on SC formation under physiological conditions

To obtain a detailed characterization of MRC organization in D. melanogaster, we isolated mitochondria from wild-type adults and, after solubilization with digitonin, we performed Blue-Native Gel Electrophoresis (BNGE) followed by mass spectrometry analysis of the gel lanes, using ‘Complexome Profiling (CP)’ 48 and thus obtaining a profile of peptide intensities from most OXPHOS subunits along the electrophoresis lane (Figure 1A). This allowed us to unequivocally determine the identity of the main bands that can be visualized by Coomassie staining of BNGE gels (Figure 1B). The identity of the bands corresponding to complex I (CI) and complex IV (CIV) was also confirmed using specific in-gel activity (IGA) assays (Figure 1B). These analyses verified that D. melanogaster mitochondria contain extremely low amounts of high-molecular weight CI-containing SCs 46,47, using the same solubilization and electrophoresis conditions in which the SCs are readily detectable in mammalian cells and tissues 29,43,49. As previously described 47, dimeric complex V (CV2) is easily visualized by BNGE and present in similar amounts as monomeric CV in D. melanogaster mitochondrial membranes solubilized with digitonin (Figure 1A, B). This CV2 species is not a strongly bound dimer, as it disappears when the samples are solubilized using a harsher detergent such as dodecylmaltoside (DDM) 47 (Figure 1C). Conversely, CI is mainly found as a free complex in the native gels irrespective of whether the mitochondria had been solubilized with digitonin or DDM (Figure 1A, B and C). The other minor CI-containing band corresponding to the fraction associated with CIII2, accounts for about 3% of the total amounts of CI and CIII2, according to label-free mass spectrometry quantifications in the CP analysis and densitometric quantifications of the CI-IGA signals (Figure 1A and Figure S1A). CIV activity is absent in this band both in digitonin- and DDM-solubilized samples (Figure 1B, C), whereas it is present in the bands that correspond to individual CIV and dimeric CIV (CIV2) detected both in digitonin-and DDM-treated samples, as well as in SC III2IV1, which is present only in the digitonin-solubilized samples. This is different from mammalian mitochondria in which SC III2IV1 is present also in DDM-solubilized mitochondria, probably due to the tight binding of CIII2 to CIV through COX7A2L/SCAF1 4,29,50. The latter does not have a homolog in D. melanogaster even though this species has three different COX7A isoforms (named COX7A, COX7AL and COX7AL2) that exhibit a tissue-specific expression pattern. CP analysis of Drosophila mitochondria only detected COX7A (mammalian COX7A1 homolog) and COX7AL2 (mammalian COX7A2 homolog), whereas COX7AL, that is solely expressed in testis, was not found. Therefore, SC I1III2 can be considered the only stable SC species in physiological conditions in D. melanogaster, yet containing a minute fraction of the total CI and CIII2.

D. melanogaster mitochondrial respiratory chain does not rely on SC formation under physiological conditions

(A) Complexome profiling of wild-type D. melanogaster mitochondria. Heatmaps show relative abundance of MRC subunits belonging to complex I (CI), complex II (CII), complex III2 (CIII) and complex IV (CIV). Color scale of normalized peptide intensities are 0 (black), 96th percentile (yellow) and 1 (red). (B) BN-PAGE separation of mitochondria from wild-type D. melanogaster solubilized with digitonin. Native gels were either stained with Coomassie R250 or analyzed by in-gel activity (IGA) for complex I (CI) and complex IV (CIV). (C) BN-PAGE separation of mitochondria from wild-type D. melanogaster solubilized with dodecylmaltoside (DDM). Native gels were either stained with Coomassie R250 or analyzed by in-gel activity assay (IGA) for complex I (CI) and complex IV (CIV).

Perturbations of CIV assembly result in increased formation of SC I1III2

COA8 is a CIV assembly factor the defects of which cause isolated mitochondrial CIV deficiency in human and mouse 51,52, as well as in Drosophila melanogaster 53,54. Consistent with the role of Coa8 in CIV biogenesis, CP analysis of mitochondria from Coa8 knockout (Coa8KO) flies showed a clear decrease in fully assembled CIV and in all the CIV-containing species (Figure 2A, B) when compared to the corresponding wild-type (WT) individuals (Figure 1A and Figure 2B). Curiously, CP also showed that in the Coa8KO mitochondria the amounts of SC I1III2 were noticeably increased (Figure 2A, B). In these samples, complexes I and III2 build a stable SC species containing ∼16% of the total amount of CI, as visualized by CI-IGA, as well as by western blot (WB) and immunodetection of specific CI and CIII2 subunits after BNGE in DDM-solubilized mitochondria (Figure 2C, D, Figure S1A and S2A). We initially speculated that the ∼5-fold increase of SC I1III2 formation could be linked to the release of III2 from SC III2IV1 induced by the strong reduction in CIV amounts when Coa8 is absent.

Severely perturbed CIV assembly results in increased formation of SC I1III2

(A) Complexome profiling of Coa8 KO D. melanogaster mitochondria. Heatmaps show relative abundance of MRC subunits belonging to complex I (CI), complex II (CII), complex III2 (CIII) and complex IV (CIV). Color scale of normalized peptide intensities are 0 (black), 96° percentile (yellow) and 1 (red). (B) Average MS profiles depicted as relative abundance of MRC enzymes in natively separated complexes from wild-type (top) and Coa8 KO (bottom) fly mitochondria. Profiles of complexes I, III2 and V (CI, CIII and CV) are plotted as average peptide intensity of the specific subunits identified by MS for each complex vs. apparent molecular weight. The increase in the relative abundance of SC I1III2 in Coa8 KO mitochondria is indicated by a black arrow. (C) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from wild-type (w1118) and Coa8 KO (Coa8KO) flies. (D) BN-PAGE, western blot immunodetection of MRC complexes from wild-type (w1118) and Coa8 KO (Coa8KO) fly mitochondria using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II).

To test this hypothesis, we modulated Coa8 expression via UAS/GAL4 system using RNAi driven by a ‘mild’ ubiquitous GAL4 driver (da-gal4). With this system, the Coa8 mRNA levels were reduced to ∼60% of the control (Figure 3A). However, these flies showed comparable levels of fully assembled CIV (Figure 3B, Figure S2B). Interestingly, in this case there was also an increased formation of SC I1III2 from ∼3% in the control to ∼10% in the mild Coa8RNAi (Figure 3B, C, Figure S1B). Therefore, both strong and weak perturbations of CIV assembly produce an increased formation of CI-containing SCs in D. melanogaster, irrespective of whether they result in CIV deficiency or not.

Mildly perturbed CIV assembly results in increased formation of SC I1III2

(A) Relative quantification (RQ) of Coa8 mRNA expression in control (da-gal4>+) and Coa8 KD (da-gal4>Coa8RNAi) flies measured by qPCR. Data are plotted as mean ± SD (n = 3 biological replicates, Student’s t test *p ≤ 0.05). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Coa8 KD (da-gal4>Coa8RNAi) flies. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Coa8 KD (da-gal4>Coa8RNAi) fly mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II).

Enhanced formation of SC I1III2 does not result in increased respiratory rates

SC formation was proposed to serve as a means to favor electron transfer between the complexes and therefore increase the efficiency of CI-fueled respiration 1. With this in mind, the complete Coa8KOand mild Coa8RNAi fly mitochondria, which show increased amounts of SC I1III2 compared to the WT controls, provide an excellent opportunity to test this possibility. Oxygen consumption activities of fly homogenates in presence of different substrates and inhibitors were analyzed by high-resolution respirometry (Figure 4A, B). The significant decrease in CIV enzymatic activity in the Coa8KO (Figure 4C) was not reflected by reduced oxygen consumption (Figure 4A). This could be explained as a result of a high CIV excess in fly mitochondria, in which the observed 60% reduction in CIV enzyme activity is still above the threshold at which the CIV defect determines lower respiratory rates 55,56. In contrast, the mild reduction in Coa8 mRNA levels did not result in CIV enzymatic deficiency but produced a slight elevation in CI amounts and activity (Figure S1B, Figure 4D), which could be related to the increased SC I1III2 formation. However, the Coa8-KD mitochondria did not show any differences in respiration with either CI-linked or CII-linked substrates. Also, the increased and stable interactions between complexes I and III2 in the Coa8 deficient models did not produce a preferential utilization of electrons coming from CI, which would be the prediction if SC formation increased electron transfer efficiency 13.

Enhanced formation of SC I1III2 does not result in increased respiration

(A-B) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on (A) Coa8 KO and (B) Coa8 KD fly homogenates compared to relative controls. Data are plotted as mean ± SD (n = 4 biological replicates). (C-D) Kinetic enzyme activity of individual MRC complexes in (C) Coa8 KO and (D) Coa8 KD compared with the relative control individuals, normalized by citrate synthase (CS) activity. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, **p ≤ 0.01, ****p ≤ 0.0001).

Mild perturbation of CIII2 biogenesis also enhances SC formation in D. melanogaster

To determine whether increased SC I1III2 formation was specific for CIV deficient flies, we targeted CIII2 by knocking down the expression of Bcs1. BCS1L, the human homolog, is fundamental for a correct CIII2 biogenesis, being responsible for the incorporation of the catalytic subunit UQCRFS1 in the last steps of CIII2 maturation 57,58. To obtain a severe CIII2 defect in D. melanogaster, we crossed a ‘strong’ ubiquitous GAL4 driver (act5c-gal4) line with a UAS-Bcs1 RNAi responder line 59. The knockdown efficiency was high, with a ∼75% decrease in Bcs1 expression at the mRNA level (Figure 5A). In this model, D. melanogaster development was severely impaired causing an arrest at the larval stage 59. The strong Bcs1RNAi caused also a significant decrease in fully assembled CIII2 levels (Figure 5B, C, Figure S2C) and in CIII2 enzymatic activity of about 50% (Figure 5D). Consistent with the observed CIII2 deficiency, both the CI- and CII-linked respiration rates were significantly decreased by around 40% (Figure 5E).

Mild perturbation of CIII2 biogenesis enhances SC formation in D. melanogaster.

(A) Relative quantification (RQ) of Bcs1 mRNA expression in control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae measured by qPCR. Data are plotted as mean ± SD (n = 3 biological replicates, Student’s t test ***p ≤ 0.001). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (act5c-gal4>+) and three Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (D) Kinetic enzyme activity of individual MRC complexes in control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, **p ≤ 0.01, ****p ≤ 0.0001). (E) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) homogenates compared to relative controls. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p ≤ 0.0001). (F) Relative quantification (RQ) of Bcs1 mRNA expression in control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae measured by qPCR. Data are plotted as mean ± SD (n = 3 biological replicates, Student’s t test ***p ≤ 0.001). (G) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae. (H) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Bcs1 KD (da-gal4>Bcs1RNAi) larvae mitochondria samples using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (I) Kinetic enzyme activity of individual MRC complexes in control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, **p ≤ 0.01). (J) High-resolution respirometric (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR - pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) homogenates compared to relative controls. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p ≤ 0.0001).

In contrast, less pronounced decreases in Bcs1 expression (Figure 5F) by using the mild da-gal4 driver instead, did not produce a noticeable CIII2 enzymatic defect (Figure 5I). However, the mild Bcs1-KD mitochondria showed a very different pattern of CI distribution than the controls (Figure 5G, H, Figure S1D), showing much higher levels of SC I1III2 as well as the appearance of a new higher molecular weight band containing also CIV, as revealed by WB and immunodetection analyses (Figure 5G, H), similarly to the mammalian respirasome (SC I1III2IV1). At the functional level, this was associated with a ∼1.5-fold increase in CI enzymatic activity (Figure 5I), which is proportional to the increase in total CI amounts (Figure S1D), and in higher CI-linked respiration rates but only by ∼1.2-fold (Figure 5J). CII-linked respiration was the same in the mild Bcs1RNAi samples as in the controls. Therefore, the formation of respirasome-like SCs in these mitochondria did neither increase the efficiency of electron transfer from CI nor determine a diversion of the electronic routes giving preference to the SC-bound CI.

Mild perturbation of CI biogenesis also leads to increased SC assembly

To understand the effect of the strong and mild perturbations in CI biogenesis on MRC organization in D. melanogaster, we employed a similar strategy as that for CIII (see above). Crossing the strong ubiquitous act5c-gal4 driver fly line with the UAS-Ndufs4 RNAi responder line, produced a decrease in Ndufs4 mRNA expression of ∼90% (Figure 6A). Defects in NDUFS4 are a major cause of CI deficiency-associated mitochondrial disease in humans 60 and the mouse and D. melanogaster animal models display CI deficiency and pathological phenotypes 61,62. Accordingly, the strong reduction in Ndufs4 expression observed in our models resulted in developmental arrest and a significant decrease in fully assembled CI levels by ∼40% (Figure 6B, C, Figure S1E) and in a proportional decrease in NADH:CoQ oxidoreductase enzyme activity in the larvae (Figure 6D). BNGE analysis of the strong Ndufs4RNAi D. melanogaster mitochondria, revealed the presence of a CI subassembly, containing the core Ndufs3 subunit (Figure 6C, Figure S2E) but lacking NADH-dehydrogenase activity (Figure 6B). This is similar to what is observed in NDUFS4-deficient human and mouse, which accumulate the so-called ∼830 kDa intermediate lacking the N-module and stabilized by the NDUFAF2 assembly factor 63–65. In contrast, CIV levels and enzyme activity were significantly increased by 1.5-fold in the strong Ndufs4-KD (Figure 6C, D, Figure S2E). CI-linked respiration measured in isolated mitochondria from these flies was significantly lower than in the controls, whereas the CII-linked respiration was comparable to the control (Figure 6E). These observations are compatible with the isolated CI defect displayed by the strong Ndufs4-KD flies.

Mild perturbation of CI biogenesis enhances SC formation in D. melanogaster.

(A) Relative quantification (RQ) of Ndufs4 mRNA expression in control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae measured by qPCR. Data are plotted as mean ± SD (n = 3 biological replicates, Student’s t test **p ≤ 0.01). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (act5c-gal4>+) and three Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (D) Kinetic enzyme activity of individual MRC complexes in control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, **p ≤ 0.01, ***p ≤ 0.001). (E) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) homogenates compared to relative controls. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p ≤ 0.0001). (F) Relative quantification (RQ) of Ndufs4 mRNA expression in control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae measured by qPCR. Data are plotted as mean ± SD (n = 3 biological replicates, Student’s t test ***p ≤ 0.001). (G) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae. (H) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (I) Kinetic enzyme activity of individual MRC complexes in control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean ± SD (n = 3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, *p ≤ 0.05). (J) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR - pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>RNAi) homogenates compared to relative controls. Data are plotted as mean ± SD (n = 4 biological replicates, two-way ANOVA with Sidak’s multiple comparisons).

Conversely, when Ndufs4 mRNA expression was reduced to about half of the control levels (Figure 6F), a milder defect in CI abundance (Figure S1F, Figure S2F), which did not affect either NADH:CoQ oxidoreductase enzymatic activity or respiratory capacity was observed (Figure 6G, H, I, J) in the adult flies. However, smaller amounts of the inactive sub-CI were still detectable (Figure 6H). Interestingly enough, this milder perturbation of CI biogenesis by reducing the amounts of Ndufs4 also produced an increase in the formation of SC I1III2 (Figure 6G, H), without any changes in respiratory performance compared to the controls (Figure 6J).

Discussion

The first description of MRC SCs in the early 2000’s led to opposite opinions on whether these were real and functionally relevant entities. On the one hand, researchers argued that the random collision model and diffusion of the individual MRC complexes was well established experimentally and the idea of the SCs did not fit with these observations. On the other hand, others considered that these SCs were real entities and therefore they must have a functional relevance, mainly as a means to enhance electron transfer between the individual complexes. Presently, the existence of the SCs is not debated anymore, especially after the determination of the high resolution structures by cryo-electron microscopy (EM), first of the mammalian respirasomes (reviewed in 66), followed by that of other mammalian SCs 3,4, and of mitochondrial respiratory chain SCs from other eukaryotic species 39–42,67,68. However, whether SCs provide any catalytical advantage to the MRC or not, is still being debated and opposing views continue to exist 18,69–72. The recent resolution of MRC SCs from different eukaryotic species has revealed that the relative arrangement of the complexes within the SCs and the bridging subunits varies substantially depending on the species. Therefore, given the conservation of the structures of the individual complexes, one could argue that if SC formation was of capital importance for MRC function, the way the complexes interact should be strictly conserved as well. Importantly, neither in the mammalian respirasomes, nor in the mammalian and plant SC I1III2 there is any evidence of substrate channeling, as the CoQ binding sites in CI and CIII2 are too far apart to allow for this phenomenon to happen 3,4,18,41. This is in agreement with different sets of functional data indicating that CoQ is interchangeable between the CI-containing SCs and the rest of the MRC 24,25,29,37. Therefore, this contrasts with the possible segmentation of the CoQ pool - one dedicated to the respirasome and the other to the FADH2-linked enzymes – which has been proposed to take place as a consequence of respirasome formation 12–14. A further element against the notion that SC formation is essential for mitochondrial function, is the fact that in normal conditions the MRC of D. melanogaster is predominantly organized based on individual complexes, as shown by us in this work and by others 46,47. The Drosophila organization was justified by a tighter packing within the mitochondrial cristae and a higher concentration of the mobile electron transporters, as a way to compensate for the lack of SCs 47. However, inter-species differences in MRC organization do not appear to be of great relevance to determine the level of functionality 28. For example, disaggregation of CI from the SCs in A. thaliana, where there are normally present, did not affect plant viability 73. Altogether, these observations argue against the strict requirement of SC formation to maintain MRC function.

The second main proposed role for SC is that of stabilizing and/or favoring the assembly of the individual complexes, especially that of CI 37,38,74. Very recently, the D. melanogaster CI structure has been solved 75,76. This has provided a structural explanation as to why this complex is majorly found as an individual entity, which seems to be due to a more stable association of the NDUFA11 subunit in the D. melanogaster complex 76. Subunit NDUFA11 stabilizes the transmembrane helix anchoring the lateral helix of subunit MT-ND5, bridging the two parts of the membrane arm of CI. In other systems, CIII2 and, thus, SC formation seem to be important for stabilizing the interaction of NDUFA11 in the CI membrane domain instead 3,5,41,77. However, in this work we show that there are ways to induce the formation of CI-containing SCs in fruit fly mitochondria where they normally do not exist. For example, whereas a strong impairment of CIII2 biogenesis resulted in CIII2 deficiency, decreased respiration and equal amounts of assembled CI, a mild perturbation of CIII2 biogenesis did not result in any MRC deficiency but rather in increased CI total amounts, mainly due to enhanced SC formation. Similarly, strong decreases in Ndufs4 expression resulted in low CI amounts and activity, but milder decreases did not affect CI function although CI was redistributed into supramolecular species. In contrast, both the complete KO and mild KD of Coa8, induced a significantly increased formation of CI-containing SCs. Even though the Coa8KO flies display a significant reduction in CIV amounts and activity, the loss of COA8 is associated with milder phenotypes in humans, mouse and flies compared with the lack or dysfunction of other CIV assembly factors 51,52,54. Therefore, we propose that a partial loss of CIV assembly also induces the formation of SCs in D. melanogaster. These observations are in line with the idea that in situations of suboptimal MRC complex biogenesis, formation of CI-containing SCs could be a way to structurally stabilize the system and preserve its function 65,78. In the context of the cooperative assembly model, where partially assembled complexes get together before completion forming SC precursors 38, one could envision that slower assembly kinetics would increase the chance of interactions at early stages, letting SC assembly occur in a stable way. Difference in the kinetics of CIV assembly have been observed between human and mouse fibroblasts, which is slower in the human cells 79. Therefore, differences in assembly kinetics could also explain the observed different amounts and stoichiometries of the MRC SCs in different organisms. This is an interesting open question that will deserve further future investigation. In any case, this enhanced SC formation did not translate in an increase in respiratory function nor in a change in substrate preference in any of the tested models. If CI-containing SC formation enhanced respiratory activity significantly, we should have had detected a noticeable increase in CI-linked respiration in all the models of mild perturbation of CIV, CIII2 and CI. Although we did observe higher respiration with CI-linked substrates in the mild Bcs1-KD, this was even lower to the increase in CI enzymatic activity and total abundance.

Therefore, we conclude that the main role of SC formation is to provide structural stability to the MRC, principally for CI, rather than to enhance electron transfer between the complexes during respiration.

Key resources table

Method details

Fly stocks and maintenance

Fly stocks were raised on standard cornmeal medium and kept at 23°C, 70% humidity on a 12:12 hours light/dark cycle. Strains used in this study were obtained from Bloomington Drosophila Stock Center (BDSC) and Vienna Drosophila Resource Center (VDRC). Genotypes used in this study were: act5c-gal4 (BDSC 4414), da-gal4 (BDSC 8641), UAS-Ndufs4 RNAi (VDRC 101489), UAS-Bcs1 RNAi (BDSC 51863), UAS-Coa8 RNAi (VDRC 100605). Control strains were obtained in each experiment by crossing the specific gal4 driver line with the genetic background flies w1118. Coa8 KO flies were generated by Wellgenetics Inc. by using CRISPR/Cas9 technology, generating a 676bp deletion, from the −49th nucleotide relative to ATG to the −34th nucleotide relative to the stop codon of Coa8.

RNA isolation, reverse transcription, and qRT-PCR

Total RNA was extracted from 10 individuals for each genotype using TRIzol method (Thermo Fisher Scientific) according to the manufacturer’s protocol. Reverse transcription was performed with the GoScript Reverse Transcriptase kit (Promega). qRT-PCRs were performed using GoTaq qPCR SYBR Green chemistry (Promega) and a Bio-Rad CFX 96 Touch System (Bio-Rad). The 2-ΔΔCt method was used to calculate the expression levels of the targets (Bcs1, Ndufs4, Coa8) using Rp49 as endogenous control. The oligonucleotides used are listed in the Key Resource Table

Isolation of Mitochondria

Mitochondria from D. melanogaster larvae and adults were prepared by homogenization and differential centrifugation as described in 81. Protein concentration of mitochondrial extracts was measured with the Bio-Rad protein assay, based on the Bradford method.

Blue-native polyacrylamide gel electrophoresis (BN-PAGE) and in-gel activity (IGA) assays

Isolated mitochondria were solubilized in 1.5M aminocaproic acid, 50 mM Bis-Tris/HCl pH 7.0. The samples were solubilized with 4 mg of digitonin (Calbiochem) or 4 mg n-dodecyl β-D-maltoside (Sigma) per mg of protein. After 5 min of incubation on ice, samples were centrifuged at 18,000 X g at 4 °C for 30 min. The supernatant was collected and resuspended with Sample Buffer (750 mM aminocaproic acid, 50 mM Bis-Tris/HCl pH 7.0, 0.5 mM EDTA and 5% Serva Blue G). Native samples were separated using NativePAGE 3-12% Bis-Tris gels (Thermo Fisher Scientific) according to the manufacturer’s protocol. For Coomassie staining, gels were stained with Coomassie R 250 for 20 minutes and destained/fixed using 20% methanol, 7% acetic acid. For in-gel activity assays, gels were stained with the following solutions: complex II (succinate dehydrogenase): 5 mM Tris–HCl pH 7.4, 0.2 mM phenazine methosulfate (Sigma), 20 mM succinate, and 1 mg/ml nitrotetrazolium blue chloride; Complex IV (cytochrome c oxidase): 50 mM potassium phosphate pH 7.4, 1 mg/ml 3′,3′-diaminobenzidine tetrahydrochloride hydrate (Sigma), 24 units/ml catalase from bovine liver (Sigma), 1 mg/ml cytochrome c from equine heart (Sigma), and 75 mg/ml sucrose 82.

Complexome profiling

Mitochondria from D. melanogaster were analyzed by complexome profiling 83. Isolated mitochondria (0.2 mg) were solubilized with 6 g digitonin/g protein in 50 mM NaCl, 5 mM 6-aminohexanoic acid, 1 mM EDTA, 50 mM imidazole/HCl, pH 7.0. After centrifugation at 22,000 X g for 20 min at 4°C, the supernatant was supplemented with Coomassie brilliant blue G250 and proteins were separated by 4-16% gradient BN-PAGE. Digitonin-solubilized mitochondrial proteins from bovine heart were loaded as molecular mass standards. Gel lanes were cut into 60 slices, transferred to a 96-well filter microtiter plate (Millipore), and destained in 50% (v/v) methanol, 50 mM ammonium bicarbonate. After destaining, in-gel digestion with trypsin was performed. Tryptic peptides were separated by liquid chromatography and analyzed by tandem mass spectrometry (LC-MS/MS) in a Q-Exactive 2.0 Orbitrap Mass Spectrometer (2.8 SP1) equipped with an Easy nLC1000 nano-flow ultra-high-pressure liquid chromatography system (Thermo Fisher Scientific) at the front end. Thermo Scientific Xcalibur 3.1 Software Package was used for data recording. MS RAW data files were analyzed using MaxQuant (version 1.5.0.25). The extracted spectra were matched against the Drosophila melanogaster Uniprot Reference Sequence database (release 2020_04). Database searches were done with 20 ppm match tolerances. Trypsin was selected as the protease with two missed cleavages allowed.

Dynamic modifications included N-terminal acetylation and oxidation of methionine. Cysteine carbamidomethylation was set as a fixed modification. Keratins, and trypsin were removed from the list. The abundance of each protein was determined by label-free quantification using the composite intensity based absolute quantification (iBAQ) values determined by MaxQuant analysis and was corrected for loading and MS sensitivity variations between samples based on the total iBAQ value for all detected complex V subunits. Gel migration profiles were created for each protein and normalized to the maximum abundance. Profiles of the identified mitochondrial proteins were hierarchically clustered by distance measures based on Pearson correlation coefficient (uncentered) and the average linkage method using the NOVA software package v0.5 84. The visualization and analysis of the heatmaps representing the normalized abundance in each gel slice by a three-color code gradient (black/yellow/red) were done using Microsoft Excel 2019 and Graph Pad Prism 8.4.3. The mass calibration for the BN-PAGE gels was performed as previously described 85. Membrane proteins were calibrated using the well-known molecular masses of respiratory chain complexes and supercomplexes from bovine heart mitochondria. The soluble proteins were, however, calibrated using the following set of Drosophila proteins: ATPB (51 kDa), malate dehydrogenase (72 kDa, dimer), citrate synthase (100 kDa, dimer), ETFA/B (122, heterodimer), heat shock protein 60 (410 kDa, heptamer), ALDH7A1 (675 kDa, dodecamer).

Western blot and immunodetection

BN-PAGE gels were transferred to PVDF membranes in Dunn carbonate buffer (10 mM NaHCO3, 3 mM Na2CO3) applying a constant current of 300 mA at 4°C for 1 hour using a Mini Trans-Blot® Cell (Bio-Rad). For the immunodetection of specific protein targets, blotted PVDF membranes were blocked in 5% skimmed milk in PBS-T (0.1% Tween-20) at room temperature for 1 hour and then incubated overnight with primary antibodies diluted in 3% BSA in PBS-T overnight at 4°C. PVDF membranes were washed three times with PBS-T for 10 minutes, incubated with the secondary HRP-conjugated antibody for 1 hour at room temperature and washed three times with PBS-T for 10 minutes. Chemiluminescent signals were recorded using an Alliance Mini HD9 (UVITEC). Antibodies used are listed in the Key Resource Table. The primary antibodies against D. melanogaster UQCR-C2 and SdhA were a kind gift of Dr. Edward Owusu-Ansah (Columbia University, NY).

High-resolution respirometry

To measure oxygen consumption individuals were homogenized on ice in respiration buffer (120 mM sucrose, 50 mM KCl, 20 mM Tris-HCl, 4 mM KH2PO4, 2 mM MgCl2, 1 mM EGTA, 1% fatty acid-free BSA, pH 7.2). Homogenates were loaded in the chamber of an O2k-HRR (High Resolution Respirometer, Oroboros Instruments) Complex I-linked respiration was measured at saturating concentrations of malate (2 mM), glutamate (10 mM), proline (10 mM) and ADP (2.5 mM). Afterwards, complex II-linked respiration was assessed using 10 mM succinate to the reaction after inhibition of complex I with rotenone (1.25 µM).

Analysis of MRC enzymatic activities

The activities of mitochondrial respiratory chain complexes and citrate synthase (CS) were measured using kinetic spectrophotometric assays as described 81.

Statistical analysis

Statistical analysis was performed with GraphPad Prism Software, version 8.2.1. Statistical tests and significance are described in the figure captions.

Data availability

Complexome profiling data will be deposited in the ComplexomE profiling DAta Resource (CEDAR) repository 86. Data will be available upon manuscript acceptance.

eLife Assessment

This study presents valuable findings on the organization of respiratory chain complexes in mitochondria, but the evidence supporting the main claims is currently incomplete, as the structural reasons for and significance of supercomplex formation are unclear. The work will be of interest to the community working on mitochondrial bioenergetics.

Reviewer #1 (Public Review):

In their manuscript, Brischigliaro et al. show that the disruption of respiratory complex assembly results in Drosophila melanogaster results in the accumulation of respiratory supercomplexes. Further, they show that the change in the supercomplex abundance does not impact respiratory function suggesting that the main role of supercomplex formation is structural. Overall, the manuscript is well written and the results and conclusion are supported. The D. melanogaster system, in which the abundance of supercomplexes can be altered through the genetic disruption of the assembly of the individual complexes, will be important for the field to discover the role of the supercomplexes. This manuscript will be of broad interest to the field of mitochondrial bioenergetics. The findings are valuable and the evidence is convincing.

Strengths:

The system developed in which the relative levels of SCs can be varied will be extremely useful for studying SC physiology.

The experiments are clearly described and interpreted.

Weaknesses:

The statement in the abstract regarding low amounts of SCs in "insect tissues" needs further support or should be narrowed. I am only aware of detailed characterization of the mitochondrial SC composition from D. melanogaster, which is insufficient to make a broad statement about the large and diverse category of insects. This should be rewritten.

In the introduction (line 76) and discussion (line 283), the authors reference the CoQ binding sites in CI and CIII2 being "too far apart" to allow for substrate channeling. The distance between the active sites, though significant, is insufficient to rule out substrate channeling. A stronger argument arises from the fact that the CoQ sites of both CI and CIII2 are open to the membrane and that there are no clear barriers for the free exchange of CoQ with the membrane pool.

Line 195, the slight elevation in CI amounts referred to here, does not appear to be statistically significant and therefore should not be discussed a being altered relative to the control.

Figure 4H, the assignments of the observed larger bands seem incorrect. The largest band (currently assigned as SC I1+III2+IV1) represents too large of a shift for only the addition of CIV and the band currently assigned at SC I1+III2 appears to also contain CIV. The identity of these bands should be reevaluated and additional experiments are needed to definitively prove their identity. This uncertainty should be addressed experimentally or made more explicit in the text.

Line 302, the authors state that the structural basis for less SC in D. melanogaster is "due to a more stable association of the NDUFA11 subunit..." However, this would not result is a less stable SC association and only explains why NDUFA11 is more stably associated with CI in the absence of CIII2. The more likely structural reason for the observation of less SC in D. melanogaster is the N-terminal truncation of Dm-NDUFB4 relative to mammalian NDUFB4. This truncation results in the loss of a major SC interaction site between CI and CIII2 in the matrix.

Reviewer #2 (Public Review):

Respiratory chain complexes assemble in higher-ordered structures termed supercomplexes or respirasomes. The functional significance of these assemblies is currently investigated, there are two main hypothesis tested, namely that supercomplexes provide kinetic advantages or structural stability. Here, the authors use the fruitfly to reveal that, while the respiratoy chain in the organism normally does not form higher-order assemblies, it does so under conditions when their assembly is impaired. Because the rather moderate increase in supercomplex formation does not change oxygen consumption stimulated by CI or CII substrate, the authors conclude that supercomplex formation has more a structural than a functional role. The main strength of this work is that the technical quality of the experiments is high and that the authors induced defects in respiratory chain assembly through sets of well-controlled genetic models. The obtained data are mostly descriptive using standard approaches and are very well executed. The authors claim that their experiments allow to conclude that the role of supercomplex formation is restricted to a structural role and, hence, exclude a function directly related to electron transport efficiency. However, while the authors can show convincingly that supercomplexes form in the mutants, but not in the wild type, their main claim is not well supported by data and both the structural mechanism of supercompelx formation and their significance remain unknown. While the supercomplex formation observed only in mitochondrial mutants per se is interesting, it would be good to great to define structural aspects of supercomplex formation and their potential impact on the stability of the respiratory chain complexes in these mutants.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

In their manuscript, Brischigliaro et al. show that the disruption of respiratory complex assembly results in Drosophila melanogaster results in the accumulation of respiratory supercomplexes. Further, they show that the change in the supercomplex abundance does not impact respiratory function suggesting that the main role of supercomplex formation is structural. Overall, the manuscript is well written and the results and conclusion are supported. The D. melanogaster system, in which the abundance of supercomplexes can be altered through the genetic disruption of the assembly of the individual complexes, will be important for the field to discover the role of the supercomplexes. This manuscript will be of broad interest to the field of mitochondrial bioenergetics. The findings are valuable and the evidence is convincing.

Strengths

The system developed in which the relative levels of SCs can be varied will be extremely useful for studying SC physiology.

The experiments are clearly described and interpreted.

Weaknesses

The statement in the abstract regarding low amounts of SCs in "insect tissues" needs further support or should be narrowed. I am only aware of detailed characterization of the mitochondrial SC composition from D. melanogaster, which is insufficient to make a broad statement about the large and diverse category of insects. This should be rewritten.

Thank you for the comment. We have amended the text accordingly.

In the introduction (line 76) and discussion (line 283), the authors reference the CoQ binding sites in CI and CIII2 being "too far apart" to allow for substrate channeling. The distance between the active sites, though significant, is insufficient to rule out substrate channeling. A stronger argument arises from the fact that the CoQ sites of both CI and CIII2 are open to the membrane and that there are no clear barriers for the free exchange of CoQ with the membrane pool.

Thank you for the comment. We have modified both sentences accordingly.

Line 195, the slight elevation in CI amounts referred to here, does not appear to be statistically significant and therefore should not be discussed a being altered relative to the control.

To address this point of criticism we have revisited the statistical analysis, originally done by 2-way ANOVA and post-hoc test. After giving it some thought, we now consider that this might not have been the correct way to analyze either the mitochondrial respiratory chain (MRC) activity data or the densitometric quantifications. We have now used unpaired two-tailed Student’s t-test to compare the pairs of either KO or KD vs CTRL. The reason is that since the measurement of each individual MRC activity is actually an independent assay, it should be considered separately. The same applies to the densitometry because the absolute values of the intensity of individual CI and that within SCs largely differ. Therefore, we think that it is more correct to compare the abundance of individual CI in the WT vs. either KO or KD pairs and the abundance of the CI in SC independently using a t-test. With these new statistical analyses, the difference in the enzyme activity of CI reported in figure 4D is now significant, which we consider reflects better our observations. Also, with these new analyses, the difference in the amounts of CI+CIII are significantly higher in the Coa8 KD (Figure S1B). Therefore, the original affirmation is correct and we have left the sentence as it was.

Figure 4H, the assignments of the observed larger bands seem incorrect. The largest band (currently assigned as SC I1+III2+IV1) represents too large of a shift for only the addition of CIV and the band currently assigned at SC I1+III2 appears to also contain CIV. The identity of these bands should be reevaluated and additional experiments are needed to definitively prove their identity. This uncertainty should be addressed experimentally or made more explicit in the text.

Thank you for the comment. Taking a closer look at the images, we have to agree with the Reviewer that the assignment was incorrect. The higher band is too large indeed and the reviewer is correct that the band that we previously assigned as CI1+CIII2 does appear to contain CIV as well. Therefore, we have changed the labeling of that to CI1+CIII2+CIV1 because the stoichiometry is compatible with the apparent MW. Also, we have renamed the higher MW band to HMW-SC (high-MW SC) of uncertain nature (unknown stoichiometry) but clearly containing all three complexes I, III and IV. We amended the text (lines 219-221) plus figures 5H and S1 accordingly.

Line 302, the authors state that the structural basis for less SC in D. melanogaster is "due to a more stable association of the NDUFA11 subunit..." However, this would not result is a less stable SC association and only explains why NDUFA11 is more stably associated with CI in the absence of CIII2. The more likely structural reason for the observation of less SC in D. melanogaster is the N-terminal truncation of Dm-NDUFB4 relative to mammalian NDUFB4. This truncation results in the loss of a major SC interaction site between CI and CIII2 in the matrix.

Thank you for pointing this out. We have amended the text accordingly.

Reviewer #2 (Public Review):

Respiratory chain complexes assemble in higher-ordered structures termed supercomplexes or respirasomes. The functional significance of these assemblies is currently investigated, there are two main hypothesis tested, namely that supercomplexes provide kinetic advantages or structural stability. Here, the authors use the fruitfly to reveal that, while the respiratoy chain in the organism normally does not form higher-order assemblies, it does so under conditions when their assembly is impaired. Because the rather moderate increase in supercomplex formation does not change oxygen consumption stimulated by CI or CII substrate, the authors conclude that supercomplex formation has more a structural than a functional role. The main strength of this work is that the technical quality of the experiments is high and that the authors induced defects in respiratory chain assembly through sets of well-controlled genetic models. The obtained data are mostly descriptive using standard approaches and are very well executed. The authors claim that their experiments allow to conclude that the role of supercomplex formation is restricted to a structural role and, hence, exclude a function directly related to electron transport efficiency. However, while the authors can show convincingly that supercomplexes form in the mutants, but not in the wild type, their main claim is not well supported by data and both the structural mechanism of supercompelx formation and their significance remain unknown. While the supercomplex formation observed only in mitochondrial mutants per se is interesting, it would be good to great to define structural aspects of supercomplex formation and their potential impact on the stability of the respiratory chain complexes in these mutants.

We thank the Reviewer for the positive assessment of our work and the suggestions to improve the manuscript.

Reviewer #1 (Recommendations For The Authors):

The sentence on line 90, which starts "This is in contrast with..." is unclear and needs to be rewritten.

Thank you. We have modified the sentence to make it clearer.

Lines 153 and 155, reference is made to tissue specific expression patterns but no literature reference is provided.

Thank you for the comment. The tissue specific expression patterns of the different isoforms are reported in the FlyBase database. We added the link to website in the text.

Line 188, "...homogenates in presence of..." should read "homogenates in the presence of..."

Thank you. Amended.

Line 336, "...lower to the increase..." should read "...lower than the increase..."

Thank you. Amended.

Reviewer #2 (Recommendations For The Authors):

In order to unravel the molecular mechanism by which supercomplexes form in the mutant, it would be important to identify the factor mediating this. Prime candidates would be additional proteins that co-purify of co-fractionate with the respiratory chain when they assemble into supercomplexes or changes in the lipid composition of the mitochondria, where cardiolipin has been shown to stabilize supercomplex formation. The inclusion and analysis of complexome data for all mutants would be excellent, plus an MS analysis of a purified supercomplex.

Thank you for the suggestion to which we completely agree. We have taken a closer look to the hierarchical clustering of peptide intensities in our complexome profiling data, which clusters the proteins according to their similarity in electrophoretic migration within the complexes. We have specifically looked for proteins in which the peptide intensity changed in a similar fashion as the complex I structural subunits. Among the four candidate proteins (Uniprot IDs Q8SXY6, Q95T19, Q9W0Y6, Q9VJQ3), only Q95T19 — Serine--tRNA synthetase-like protein Slimp is annotated as a mitochondrial protein. This protein is a Drosophila-specific paralog of the mitochondrial Serine-tRNA synthetase generated by gene duplication (PMID: 20870726), which carries out a function linking mitochondrial translation with mtDNA maintenance (PMID: 30943413). Therefore, in principle we would not consider it as a good candidate to be a ‘SC assembly factor’. The identification of factors promoting the formation of SC in Drosophila under these conditions is definitely an important point warranting future investigation.

The authors could define the stability of the respiratory chain complexes through metabolic pulse-chase labeling experiments. This could reveal that the role of supercomplex formation is indeed structural, improving stability.

We agree that this would be an important piece of information to understand the phenomenon we have observed. Unfortunately, it is technically impossible to perform metabolic labeling of mitochondrial proteins in whole flies. It would be possible to perform in organello pulse-chase labelling, however our previous experience indicates that complex I does not completely assemble de novo in isolated mitochondria (PMID: 20385768).

The authors should analyze oxygen consumption from mitochondria isolated from larvae as in the other experiments on enzyme activities or the (high-quality) BN-PAGE, and not from whole flies that are homogenized. Moreover, they need to determine the quantities of the complexes by complementary experiments (MS, Western blotting or spectroscopy).

Thank you for the comments. However, we believe that repeating the entire analyses with the larvae would not add significant information to the work and the main interpretation would not change, as the main claim of the paper is based on the data collected on adult flies. In addition, the band patterns of MRC complexes in the BNGE is the same between larvae and adults and therefore, does not depend on the developmental stage. Regarding the quantification of the complexes, we think that the data provided by using complementary approaches such as in gel activity assays (IGA), western blot (WB) and kinetic assays of MRC enzymatic activities, allowed us to confidently determine the amount of the individual complexes. Hence, we performed IGA assays and enzymatic activity assays (which reflect the amounts of fully assembled and functional complexes) in triplicate (independent samples). For the WB analyses, due to the scarcity of some of the antibodies available to detect the Dm MRC proteins, which were a kind gift of Dr. Edward Owusu-Ansah (Columbia University), we decided to pool the three independent samples of each group before running them through the Blue-Native gels. The densitometric curves of the WB bands (Figure S2) show the abundance of each individual MRC complex within the ‘free’ and SC forms. We prioritized the BN analyses over SDS-PAGE and WB analysis, as we consider that just measuring the steady-state levels of MRC subunits is not as informative, because it is possible that certain subunits are present in the mitochondrial membranes but not assembled into the final mature structures.

Can changes in Coenzyme Q levels explain the absence of a defect on electron transport? This could be determined for the mutant as well as the wild type animals.

We agree that this would be a relevant aspect to investigate. For example, determining whether lower CoQ levels are able to maintain the same respiratory activities in the models in which higher amounts of SCs are formed, as it was proposed in Shimada et al. (PMID: 29191512) would be very interesting. However, the fact that the mild KD models show no MRC enzymatic defects whatsoever (Figure 4D, Figure 5I and Figure 6I), provides the most straightforward explanation to the observed absence of respiratory defects.

Version 2

Abstract

Abstract

Mammalian mitochondrial respiratory chain (MRC) complexes are able to associate into quaternary structures named supercomplexes (SCs), which normally coexist with non bound individual complexes. The functional significance of SCs has not been fully clarified and the debate has been centered on whether or not they confer catalytic advantages to the non-bound individual complexes. Mitochondrial respiratory chain organization does not seem to be conserved in all organisms. In fact, and differently from mammalian species, mitochondria from Drosophila melanogaster tissues are characterized by low amounts of SCs, despite the high metabolic demands and MRC activity shown by these mitochondria. Here, we show that attenuating the biogenesis of individual respiratory chain complexes was accompanied by increased formation of stable SCs, which are missing in Drosophila melanogaster in physiological conditions. This phenomenon was not accompanied by an increase in mitochondrial respiratory activity. Therefore, we conclude that SC formation is necessary to stabilize the complexes in suboptimal biogenetic conditions, but not for the enhancement of respiratory chain catalysis.

Introduction

Mitochondria are the organelles providing most of the cellular energy in form of adenosine triphosphate (ATP) in aerobic eukaryotes. The molecular machinery responsible for energy transformation is the oxidative phosphorylation (OXPHOS) system, which is canonically composed of five multiprotein complexes embedded in the inner mitochondrial membrane. OXPHOS consists of two tightly regulated processes: electron transport and ATP synthesis. Electron transport takes place between complexes I-IV and two mobile electron carriers (coenzyme Q and cytochrome c). During electron transport, complexes I, III and IV pump protons from the mitochondrial matrix to the intermembrane space, generating a proton gradient that provides the protonmotive force exploited by complex V to synthesize ATP. In mammalian mitochondria, mitochondrial respiratory chain (MRC) complexes I, III and IV can interact with each other forming supramolecular structures described generally by the term ‘supercomplexes’ [1, 2]. MRC supercomplexes (SCs) can have different stoichiometries and compositions, ranging from the binding of only two complexes, such as the I1III2 and III2IV1 SCs [3, 4], to higher-order associations between complexes I, III and IV, with the SC of I1III2IV1 stoichiometry known as the ‘respirasome’ [1, 5–9]. Now that the association of individual MRC complexes into supramolecular structures is well established, with structures of several SC species being resolved, the debate is centered on what the functional significance of these structures might be.

Several possible roles have been proposed for SCs. First, it was suggested that the association between complex I (CI) and the obligate dimer of complex III (CIII2) would allow substrate channeling, sequestering a dedicated coenzyme Q (CoQ) pool and allowing a more efficient electron transfer between the two complexes, while separating this electronic route from those of FADH2-linked dehydrogenases (e.g. complex II) to the CIII2 not bound to CI [1, 10–15]. This increased efficiency would in turn decrease electron leak and, as a consequence, produce less reactive oxygen species (ROS) than the individual free complexes [16, 17]. However, the available high-resolution respirasome structures show that the distance between the CoQ binding sites in CI and CIII2 are far apart and exposed to the membrane, thus not supporting the notion of substrate channeling within the SC structure [4, 18]. In addition, exogenously added CoQ was necessary to sustain CI activity in the purified I1III2 SC [3], arguing against a tightly bound and segregated CoQ pool as a functional component of the SC. A large body of work from the late 1960’s to the 1980’s, resulted in the ‘random collision model’ to explain electron transfer between the respiratory chain complexes, and in the evidence that CoQ is present as an undifferentiated pool[19–23]. More recently, additional proof of the non compartmentalized electronic routes from CI to CIII2 and from complex II (CII) to CIII2, came from kinetic measurements in sub-mitochondrial particles. In these systems, MRC organization in SCs was conserved but b- and c-type cytochromes in CIII2 were equally accessible to CI-linked and CII-linked substrates [24], and CoQ reduced by CI in the respirasomes was able to reach and readily reduce external enzymes to the SCs [25]. In addition, growing evidence supports the notion that different MRC organizations exist in vivo, where varying proportions of SC vs. free complexes do not result in separate and distinct CI-linked and CII-linked respiratory activities [26–29]. This is in contrast with the idea that segregation into different types of SCs and in individual complexes is necessary for the functional interplay of the MRC, leading to the adaptation of the respiratory activity to different metabolic settings [13]. The physical proximity of CIII2 and CIV has also been suggested to promote faster electron transfer kinetics via cytochrome c [4, 30, 31], although this is a matter of debate as well [32, 33].

The second main explanation to justify the existence of SCs is that they play a structural function, stabilizing the individual complexes [34, 35] and/or serving as a platform for the efficient assembly of the complexes, with a special relevance for the biogenesis of mammalian CI [36–38].

Notably, the MRC structural organization, especially the stoichiometry, arrangements and stability of the SCs, may not be conserved in all eukaryotic species [39–42]. This is the case even within mammals, as human cells and tissues barely contain free CI, which is rather contained in SC I1+III2 and the respirasome [29, 37]. In contrast, other mammalian mitochondria (from bovine, ovine, rat or mouse) contain larger amounts of CI in its free form, even if the majority is still mostly in the form of SCs [2, 17, 28, 43–45]. The distribution of the MRC complexes between free complexes and SCs seems to differ even more in non-mammalian animal species. Several reptile species contain a very stable SC I+III2 that lacks CIV [28], and in Drosophila melanogaster practically all of CI is free, SCs being almost completely absent [46, 47]. However, comparative studies of MRC function in diverse animal species suggest that higher amounts and stability of the SCs do not correlate with increased respiratory activity/efficiency and/or reduced ROS production [28, 47].

Here we show that SCs can be stably formed in D. melanogaster mitochondria upon mild perturbations of individual CIV, CIII2 and CI biogenesis. This finding enabled us to test whether increased SC formation translated into enhanced respiration proficiency. However, MRC performance of fruit fly mitochondria did not change regardless of the presence or absence of SCs. These observations have led us to conclude that: 1) the efficiency in the assembly of the individual complexes is likely to be the main determinant of SC formation and 2) these supramolecular complexes play a more relevant role in maintaining the stability and/or supporting the biogenesis of the MRC than in promoting catalysis.

Results

D. melanogaster MRC organization does rely on SC formation under physiological conditions

To obtain a detailed characterization of MRC organization in D. melanogaster, we isolated mitochondria from wild-type adults and, after solubilization with digitonin, we performed Blue-Native Gel Electrophoresis (BNGE) followed by mass spectrometry analysis of the gel lanes, using ‘Complexome Profiling (CP)’ [48] and thus obtaining a profile of peptide intensities from most OXPHOS subunits along the electrophoresis lane (Figure 1A). This allowed us to unequivocally determine the identity of the main bands that can be visualized by Coomassie staining of BNGE gels (Figure 1B). The identity of the bands corresponding to complex I (CI) and complex IV (CIV) was also confirmed using specific in-gel activity (IGA) assays (Figure 1B). These analyses verified that D. melanogaster mitochondria contain extremely low amounts of high-molecular weight CI-containing SCs [46, 47], using the same solubilization and electrophoresis conditions in which the SCs are readily detectable in mammalian cells and tissues [29, 43, 49]. As previously described [47], dimeric complex V (CV2) is easily visualized by BNGE and present in similar amounts as monomeric CV in D. melanogaster mitochondrial membranes solubilized with digitonin (Figure 1A, B). This CV2 species is not a strongly bound dimer, as it disappears when the samples are solubilized using a harsher detergent such as dodecylmaltoside (DDM) [47] (Figure 1C). Conversely, CI is mainly found as a free complex in the native gels irrespective of whether the mitochondria had been solubilized with digitonin or DDM (Figure 1A, B and C). The other minor CI-containing band corresponding to the fraction associated with CIII2, accounts for about 3% of the total amounts of CI and CIII2, according to label-free mass spectrometry quantifications in the CP analysis and densitometric quantifications of the CI-IGA signals (Figure 1A and Figure S1A). CIV activity is absent in this band both in digitonin- and DDM solubilized samples (Figure 1B, C), whereas it is present in the bands that correspond to individual CIV and dimeric CIV (CIV2) detected both in digitonin- and DDM-treated samples, as well as in SC III2IV1, which is present only in the digitonin-solubilized samples. This is different from mammalian mitochondria in which SC III2IV1 is present also in DDM-solubilized mitochondria, probably due to the tight binding of CIII2 to CIV through COX7A2L/SCAF1 [4, 29, 50]. The latter does not have a homolog in D. melanogaster even though this species has three different COX7A isoforms (named COX7A, COX7AL and COX7AL2) that exhibit a tissue-specific expression pattern, according to FlyBase (www.flybase.org). CP analysis of Drosophila mitochondria only detected COX7A (mammalian COX7A1 homolog) and COX7AL2 (mammalian COX7A2 homolog), whereas COX7AL, that is solely expressed in testis, was not found. Therefore, SC I1III2 can be considered the only stable SC species in physiological conditions in D. melanogaster, yet containing a minute fraction of the total CI and CIII2.

D. melanogaster mitochondrial respiratory chain does not rely on SC formation under physiological conditions.

(A) Complexome profiling of wild-type D. melanogaster mitochondria. Heatmaps show relative abundance of MRC subunits belonging to complex I (CI), complex II (CII), complex III2 (CIII) and complex IV (CIV). Color scale of normalized peptide intensities are 0 (black), 96th percentile (yellow) and 1 (red). (B) BN-PAGE separation of mitochondria from wild-type D. melanogaster solubilized with digitonin. Native gels were either stained with Coomassie R250 or analyzed by in-gel activity (IGA) for complex I (CI) and complex IV (CIV). (C) BN-PAGE separation of mitochondria from wild-type D. melanogaster solubilized with dodecylmaltoside (DDM). Native gels were either stained with Coomassie R250 or analyzed by in-gel activity assay (IGA) for complex I (CI) and complex IV (CIV).

Perturbations of CIV assembly result in increased formation of SC I1III2

COA8 is a CIV assembly factor the defects of which cause isolated mitochondrial CIV deficiency in human and mouse [51, 52], as well as in Drosophila melanogaster [53, 54]. Consistent with the role of Coa8 in CIV biogenesis, CP analysis of mitochondria from Coa8 knockout (Coa8KO) flies showed a clear decrease in fully assembled CIV and in all the CIV containing species (Figure 2A, B) when compared to the corresponding wild-type (WT) individuals (Figure 1A and Figure 2B). Curiously, CP also showed that in the Coa8KO mitochondria the amounts of SC I1III2 were noticeably increased (Figure 2A, B). In these samples, complexes I and III2 build a stable SC species containing ∼16% of the total amount of CI, as visualized by CI-IGA, as well as by western blot (WB) and immunodetection of specific CI and CIII2 subunits after BNGE in DDM-solubilized mitochondria (Figure 2C, D, Figure S1A and S2A). We initially speculated that the ∼5-fold increase of SC I1III2 formation could be linked to the release of III2 from SC III2IV1 induced by the strong reduction in CIV amounts when Coa8 is absent.

Severely perturbed CIV assembly results in increased formation of SC I1III2.

(A) Complexome profiling of Coa8 KO D. melanogaster mitochondria. Heatmaps show relative abundance of MRC subunits belonging to complex I (CI), complex II (CII), complex III2 (CIII) and complex IV (CIV). Color scale of normalized peptide intensities are 0 (black), 96° percentile (yellow) and 1 (red). (B) Average MS profiles depicted as relative abundance of MRC enzymes in natively separated complexes from wild-type (top) and Coa8 KO (bottom) fly mitochondria. Profiles of complexes I, III2 and V (CI, CIII and CV) are plotted as average peptide intensity of the specific subunits identified by MS for each complex vs. apparent molecular weight. The increase in the relative abundance of SC I1III2 in Coa8 KO mitochondria is indicated by a black arrow. (C) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from wild-type (w1118) and Coa8 KO (Coa8KO) flies. (D) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control wild-type (w1118) and and three Coa8 KO (Coa8KO) fly mitochondria preparations, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II).

To test this hypothesis, we modulated Coa8 expression via UAS/GAL4 system using RNAi driven by a ‘mild’ ubiquitous GAL4 driver (da-gal4). With this system, the Coa8 mRNA levels were reduced to ∼60% of the control (Figure 3A). However, these flies showed comparable levels of fully assembled CIV (Figure 3B, Figure S2B). Interestingly, in this case there was also an increased formation of SC I1III2 from ∼3% in the control to ∼10% in the mild Coa8RNAi (Figure 3B, C, Figure S1B). Therefore, both strong and weak perturbations of CIV assembly produce an increased formation of CI-containing SCs in D. melanogaster, irrespective of whether they result in CIV deficiency or not.

Mildly perturbed CIV assembly results in increased formation of SC I1III2.

(A) Relative quantification (RQ) of Coa8 mRNA expression in control (da-gal4>+) and Coa8 KD (da-gal4>Coa8RNAi) flies measured by qPCR. Data are plotted as mean[±[SD (n[=[3 biological replicates, Student’s t test *p[≤[0.05). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Coa8 KD (da-gal4>Coa8RNAi) flies. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Coa8 KD (da-gal4>Coa8RNAi) fly mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti COX4 (complex IV) and anti-SDHA (complex II).

Enhanced formation of SC I1III2 does not result in increased respiratory rates

SC formation was proposed to serve as a means to favor electron transfer between the complexes and therefore increase the efficiency of CI-fueled respiration [1]. With this in mind, the complete Coa8KO and mild Coa8RNAi fly mitochondria, which show increased amounts of SC I1III2 compared to the WT controls, provide an excellent opportunity to test this possibility. Oxygen consumption activities of fly homogenates in the presence of different substrates and inhibitors were analyzed by high-resolution respirometry (Figure 4A, B). The significant decrease in CIV enzymatic activity in the Coa8KO(Figure 4C) was not reflected by reduced oxygen consumption (Figure 4A). This could be explained as a result of a high CIV excess in fly mitochondria, in which the observed 60% reduction in CIV enzyme activity is still above the threshold at which the CIV defect determines lower respiratory rates [55, 56]. In contrast, the mild reduction in Coa8 mRNA levels did not result in CIV enzymatic deficiency but produced a slight elevation in CI activity (Figure 4D), which is most likely due to the increased SC I1III2 formation (Figure 3B and C, Figure S1B). However, the Coa8-KD mitochondria did not show any differences in respiration with either CI-linked or CII-linked substrates. Also, the increased and stable interactions between complexes I and III2 in the Coa8 deficient models did not produce a preferential utilization of electrons coming from CI, which would be the prediction if SC formation increased electron transfer efficiency [13].

Enhanced formation of SC I1III2 does not result in increased respiration.

(A-B) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on (A) Coa8 KO and (B) Coa8 KD fly homogenates compared to relative controls. Data are plotted as mean[±[SD (n[=[4 biological replicates). (C-D) Kinetic enzyme activity of individual MRC complexes in (C) Coa8 KO and (D) Coa8 KD compared with the relative control individuals, normalized by citrate synthase (CS) activity. Data are plotted as mean[±[SD (n[=[3 biological replicates, pairwise comparisons by unpaired Student’s t test *p[≤[0.05, **p[≤[0.01, ****p[≤[0.0001).

Mild perturbation of CIII2 biogenesis also enhances SC formation in D. melanogaster

To determine whether increased SC I1III2 formation was specific for CIV deficient flies, we targeted CIII2 by knocking down the expression of Bcs1. BCS1L, the human homolog, is fundamental for a correct CIII2 biogenesis, being responsible for the incorporation of the catalytic subunit UQCRFS1 in the last steps of CIII2 maturation [57, 58]. To obtain a severe CIII2 defect in D. melanogaster, we crossed a ‘strong’ ubiquitous GAL4 driver (act5c-gal4) line with a UAS-Bcs1 RNAi responder line [59]. The knockdown efficiency was high, with a ∼75% decrease in Bcs1 expression at the mRNA level (Figure 5A). In this model, D. melanogaster development was severely impaired causing an arrest at the larval stage [59]. The strong Bcs1RNAi caused also a significant decrease in fully assembled CIII2 levels (Figure 5B, C, Figure S2C) and in CIII2 enzymatic activity of about 50% (Figure 5D). Consistent with the observed CIII2 deficiency, both the CI- and CII linked respiration rates were significantly decreased by around 40% (Figure 5E).

Mild perturbation of CIII2 biogenesis enhances SC formation in D. melanogaster.

(A) Relative quantification (RQ) of Bcs1 mRNA expression in control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae measured by qPCR. Data are plotted as mean[±[SD (n[=[3 biological replicates, Student’s t test ***p[≤[0.001). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (act5c-gal4>+) and three Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (D) Kinetic enzyme activity of individual MRC complexes in control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean[±[SD (n[=[3 biological replicates, pairwise comparisons by unpaired Student’s t test, *p[≤[0.05, ****p[≤[0.0001). (E) High resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) homogenates compared to relative controls. Data are plotted as mean[±[SD (n[=[3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p[≤[0.0001). (F) Relative quantification (RQ) of Bcs1 mRNA expression in control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae measured by qPCR. Data are plotted as mean[±[SD (n[=[3 biological replicates, Student’s t test ***p[≤[0.001). (G) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae. (H) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Bcs1 KD (da-gal4>Bcs1RNAi) larvae mitochondria samples using antibodies against specific subunits: anti-UQCRC2 (complex III), anti NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). HWM-SC: high molecular weight supercomplex. (I) Kinetic enzyme activity of individual MRC complexes in control (da-gal4>+) and Bcs1 KD (da-gal4>Bcs1RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean[±[SD (n[=[3 biological replicates, pairwise comparisons by unpaired Student’s t test, *p[≤[0.05). (J) High resolution respirometric (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR - pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Bcs1 KD (act5c-gal4>Bcs1RNAi) homogenates compared to relative controls. Data are plotted as mean[±[SD (n[=[3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p[≤[0.0001).

In contrast, less pronounced decreases in Bcs1 expression (Figure 5F) by using the mild da-gal4 driver instead, did not produce a noticeable CIII2 enzymatic defect (Figure 5I). However, the mild Bcs1-KD mitochondria showed a very different pattern of CI distribution than the controls (Figure 5G, H, Figure S1D), with the formation of a ‘respirasome-like’ SC I1III2IV1, and also of a higher molecular weight supercomplex (HMW-SC), of unknown stoichiometry, containing also complexes I, III2 and IV, as revealed by WB and immunodetection analyses (Figure 5G, H). At the functional level, this was associated with a ∼1.5-fold increase in CI enzymatic activity (Figure 5I), which is proportional to the increase in total CI amounts (Figure S1D), and in higher CI-linked respiration rates but only by ∼1.2-fold (Figure 5J). CII-linked respiration was the same in the mild Bcs1RNAi samples as in the controls. Therefore, the formation of respirasome-like SCs in these mitochondria did neither increase the efficiency of electron transfer from CI nor determine a diversion of the electronic routes giving preference to the SC-bound CI.

Mild perturbation of CI biogenesis also leads to increased SC assembly

To understand the effect of the strong and mild perturbations in CI biogenesis on MRC organization in D. melanogaster, we employed a similar strategy as that for CIII (see above). Crossing the strong ubiquitous act5c-gal4 driver fly line with the UAS-Ndufs4 RNAi responder line, produced a decrease in Ndufs4 mRNA expression of ∼90% (Figure 6A). Defects in NDUFS4 are a major cause of CI deficiency-associated mitochondrial disease in humans [60] and the mouse and D. melanogaster animal models display CI deficiency and pathological phenotypes [61, 62]. Accordingly, the strong reduction in Ndufs4 expression observed in our models resulted in developmental arrest and a significant decrease in fully assembled CI levels by ∼40% (Figure 6B, C, Figure S1E) and in a proportional decrease in NADH:CoQ oxidoreductase enzyme activity in the larvae (Figure 6D). BNGE analysis of the strong Ndufs4RNAi D. melanogaster mitochondria, revealed the presence of a CI subassembly, containing the core Ndufs3 subunit (Figure 6C, Figure S2E) but lacking NADH-dehydrogenase activity (Figure 6B). This is similar to what is observed in NDUFS4-deficient human and mouse, which accumulate the so-called ∼830 kDa intermediate lacking the N-module and stabilized by the NDUFAF2 assembly factor [63–65]. In contrast, CIV levels and enzyme activity were significantly increased by 1.5-fold in the strong Ndufs4-KD (Figure 6C, D, Figure S2E). CI-linked respiration measured in isolated mitochondria from these flies was significantly lower than in the controls, whereas the CII-linked respiration was comparable to the control (Figure 6E). These observations are compatible with the isolated CI defect displayed by the strong Ndufs4-KD flies.

Mild perturbation of CI biogenesis enhances SC formation in D. melanogaster.

(A) Relative quantification (RQ) of Ndufs4 mRNA expression in control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae measured by qPCR. Data are plotted as mean[±[SD (n[=[3 biological replicates, Student’s t test **p[≤[0.01). (B) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae. (C) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (act5c-gal4>+) and three Ndufs4 KD (act5c-gal4>Ndufs4RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti-COX4 (complex IV) and anti-SDHA (complex II). (D) Kinetic enzyme activity of individual MRC complexes in control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean[±[SD (n[=[3 biological replicates, pairwise comparisons by unpaired Student’s t test, **p[≤[0.01). (E) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR – pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>Ndufs4RNAi) homogenates compared to relative controls. Data are plotted as mean[±[SD (n[=[3 biological replicates, two-way ANOVA with Sidak’s multiple comparisons, ****p[≤[0.0001). (F) Relative quantification (RQ) of Ndufs4 mRNA expression in control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae measured by qPCR. Data are plotted as mean[±[SD (n[=[3 biological replicates, Student’s t test ***p[≤[0.001). (G) In gel-activity assays for MRC complex I (CI), complex II (CII) and complex IV (CIV) in DDM-solubilized mitochondria from control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae. (H) BN-PAGE, western blot immunodetection of MRC complexes from a pool of three control (da-gal4>+) and three Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae mitochondria samples, using antibodies against specific subunits: anti-UQCRC2 (complex III), anti-NDUFS3 (complex I), anti COX4 (complex IV) and anti-SDHA (complex II). (I) Kinetic enzyme activity of individual MRC complexes in control (da-gal4>+) and Ndufs4 KD (da-gal4>Ndufs4RNAi) larvae mitochondria normalized by citrate synthase (CS) activity. Data are plotted as mean[±[SD (n[=[3 biological replicates, pairwise comparisons by unpaired Student’s t test, *p[≤[0.05). (J) High-resolution respirometry (HRR) analyses of whole-fly homogenates. Respiration is represented by oxygen flux (JO2) measured by oxygen consumption rates (OCR - pmol/s*fly). OCR have been measured via substrate-driven respiration under saturating concentrations of substrates inducing either complex I (CI) or complex II (CII) -linked respiration. Rotenone was used to block CI-linked respiration before measuring CII-linked respiration. HRR was performed on control (act5c-gal4>+) and Ndufs4 KD (act5c-gal4>RNAi) homogenates compared to relative controls. Data are plotted as mean[±[SD (n[=[4 biological replicates, two-way ANOVA with Sidak’s multiple comparisons).

Conversely, when Ndufs4 mRNA expression was reduced to about half of the control levels (Figure 6F), a milder defect in CI abundance (Figure S1F, Figure S2F), which did not affect either NADH:CoQ oxidoreductase enzymatic activity or respiratory capacity was observed (Figure 6G, H, I, J) in the adult flies. However, smaller amounts of the inactive sub-CI were still detectable (Figure 6H). Interestingly enough, this milder perturbation of CI biogenesis by reducing the amounts of Ndufs4 also produced an increase in the formation of SC I1III2 (Figure 6G, H), without any changes in respiratory performance compared to the controls (Figure 6J).

Discussion

The first description of MRC SCs in the early 2000’s led to opposite opinions on whether these were real and functionally relevant entities. On the one hand, researchers argued that the random collision model and diffusion of the individual MRC complexes was well established experimentally and the idea of the SCs did not fit with these observations. On the other hand, others considered that these SCs were real entities and therefore they must have a functional relevance, mainly as a means to enhance electron transfer between the individual complexes. Presently, the existence of the SCs is not debated anymore, especially after the determination of the high resolution structures by cryo electron microscopy (EM), first of the mammalian respirasomes (reviewed in [66]), followed by that of other mammalian SCs [3, 4], and of mitochondrial respiratory chain SCs from other eukaryotic species [39–42, 67, 68]. However, whether SCs provide any catalytical advantage to the MRC or not, is still being debated and opposing views continue to exist [18, 69–72]. The recent resolution of MRC SCs from different eukaryotic species has revealed that the relative arrangement of the complexes within the SCs and the bridging subunits varies substantially depending on the species. Therefore, given the conservation of the structures of the individual complexes, one could argue that if SC formation was of capital importance for MRC function, the way the complexes interact should be strictly conserved as well. Importantly, neither in the mammalian respirasomes, nor in the mammalian and plant SC I1III2 there is any evidence of substrate channeling, as the CoQ binding sites in CI and CIII2 are far apart and exposed to the milieu, in principle allowing the free exchange of CoQ [3, 4, 18, 41]. This is in agreement with different sets of functional data indicating that CoQ is interchangeable between the CI-containing SCs and the rest of the MRC [24, 25, 29, 37]. Therefore, this contrasts with the possible segmentation of the CoQ pool - one dedicated to the respirasome and the other to the FADH2-linked enzymes – which has been proposed to take place as a consequence of respirasome formation [12–14]. A further element against the notion that SC formation is essential for mitochondrial function, is the fact that in normal conditions the MRC of D. melanogaster is predominantly organized based on individual complexes, as shown by us in this work and by others [46, 47]. The Drosophila organization was justified by a tighter packing within the mitochondrial cristae and a higher concentration of the mobile electron transporters, as a way to compensate for the lack of SCs [47]. However, inter-species differences in MRC organization do not appear to be of great relevance to determine the level of functionality [28]. For example, disaggregation of CI from the SCs in A. thaliana, where there are normally present, did not affect plant viability [73]. Altogether, these observations argue against the strict requirement of SC formation to maintain MRC function.

The second main proposed role for SC is that of stabilizing and/or favoring the assembly of the individual complexes, especially that of CI [37, 38, 74]. Very recently, the D. melanogaster CI structure has been solved [75, 76]. This has provided a structural explanation as to why this complex is majorly found as an individual entity, which seems to be due the fact that the N-terminal domain of the NDUFB4 subunit, present in mammals and responsible for the interaction with CIII2 within the SCs, is absent in the in the D. melanogaster subunit [76]. Additionally, fruit fly subunit NDUFA11 has an extended C-terminus that results in a tighter binding of the subunit to the rest of the CI membrane arm. NDUFA11 stabilizes the transmembrane helix anchoring the lateral helix of subunit MT-ND5, bridging the two parts of the membrane arm of CI in plants and mammals, SC formation seem to be important for stabilizing the interaction of NDUFA11 in the CI membrane domain instead [3, 5, 41, 76, 77]. However, in this work we show that there are ways to induce the formation of CI-containing SCs in fruit fly mitochondria where they normally do not exist. For example, whereas a strong impairment of CIII2 biogenesis resulted in CIII2 deficiency, decreased respiration and equal amounts of assembled CI, a mild perturbation of CIII2 biogenesis did not result in any MRC deficiency but rather in increased CI total amounts, mainly due to enhanced SC formation. Similarly, strong decreases in Ndufs4 expression resulted in low CI amounts and activity, but milder decreases did not affect CI function although CI was redistributed into supramolecular species. In contrast, both the complete KO and mild KD of Coa8, induced a significantly increased formation of CI-containing SCs. Even though the Coa8KO flies display a significant reduction in CIV amounts and activity, the loss of COA8 is associated with milder phenotypes in humans, mouse and flies compared with the lack or dysfunction of other CIV assembly factors [51, 52, 54]. Therefore, we propose that a partial loss of CIV assembly also induces the formation of SCs in D. melanogaster. These observations are in line with the idea that in situations of suboptimal MRC complex biogenesis, formation of CI-containing SCs could be a way to structurally stabilize the system and preserve its function [65, 78]. In the context of the cooperative assembly model, where partially assembled complexes get together before completion forming SC precursors [38], one could envision that slower assembly kinetics would increase the chance of interactions at early stages, letting SC assembly occur in a stable way. Difference in the kinetics of CIV assembly have been observed between human and mouse fibroblasts, which is slower in the human cells [79]. Therefore, differences in assembly kinetics could also explain the observed different amounts and stoichiometries of the MRC SCs in different organisms. This is an interesting open question that will deserve further future investigation. In any case, this enhanced SC formation did not translate in an increase in respiratory function nor in a change in substrate preference in any of the tested models. If CI-containing SC formation enhanced respiratory activity significantly, we should have had detected a noticeable increase in CI-linked respiration in all the models of mild perturbation of CIV, CIII2 and CI. Although we did observe higher respiration with CI-linked substrates in the mild Bcs1-KD, this was even lower than the increase in CI enzymatic activity and total abundance.

Therefore, we conclude that the main role of SC formation is to provide structural stability to the MRC, principally for CI, rather than to enhance electron transfer between the complexes during respiration.

Key resources table

Method details

Fly stocks and maintenance

Fly stocks were raised on standard cornmeal medium and kept at 23°C, 70% humidity on a 12:12 hours light/dark cycle. Strains used in this study were obtained from Bloomington Drosophila Stock Center (BDSC) and Vienna Drosophila Resource Center (VDRC). Genotypes used in this study were: act5c-gal4 (BDSC 4414), da-gal4 (BDSC 8641), UAS-Ndufs4 RNAi (VDRC 101489), UAS-Bcs1 RNAi (BDSC 51863), UAS-Coa8 RNAi (VDRC 100605). Control strains were obtained in each experiment by crossing the specific gal4 driver line with the genetic background flies w1118. Coa8 KO flies were generated by Wellgenetics Inc. by using CRISPR/Cas9 technology, generating a 676bp deletion, from the -49th nucleotide relative to ATG to the -34th nucleotide relative to the stop codon of Coa8.

RNA isolation, reverse transcription, and qRT-PCR

Total RNA was extracted from 10 individuals for each genotype using TRIzol method (Thermo Fisher Scientific) according to the manufacturer’s protocol. Reverse transcription was performed with the GoScript Reverse Transcriptase kit (Promega). qRT-PCRs were performed using GoTaq qPCR SYBR Green chemistry (Promega) and a Bio-Rad CFX 96 Touch System (Bio-Rad). The 2-ΔΔCt method was used to calculate the expression levels of the targets (Bcs1, Ndufs4, Coa8) using Rp49 as endogenous control. The oligonucleotides used are listed in the Key Resource Table

Isolation of Mitochondria

Mitochondria from D. melanogaster larvae and adults were prepared by homogenization and differential centrifugation as described in [81]. Protein concentration of mitochondrial extracts was measured with the Bio-Rad protein assay, based on the Bradford method.

Blue-native polyacrylamide gel electrophoresis (BN-PAGE) and in-gel activity (IGA) assays

Isolated mitochondria were solubilized in 1.5M aminocaproic acid, 50 mM Bis-Tris/HCl pH 7.0. The samples were solubilized with 4 mg of digitonin (Calbiochem) or 4 mg n-dodecyl β-D-maltoside (Sigma) per mg of protein. After 5 min of incubation on ice, samples were centrifuged at 18,000 X g at 4 °C for 30 min. The supernatant was collected and resuspended with Sample Buffer (750 mM aminocaproic acid, 50 mM Bis-Tris/HCl pH 7.0, 0.5 mM EDTA and 5% Serva Blue G). Native samples were separated using NativePAGE 3-12% Bis-Tris gels (Thermo Fisher Scientific) according to the manufacturer’s protocol. For Coomassie staining, gels were stained with Coomassie R 250 for 20 minutes and destained/fixed using 20% methanol, 7% acetic acid. For in-gel activity assays, gels were stained with the following solutions: complex II (succinate dehydrogenase): 5 mM Tris–HCl pH 7.4, 0.2 mM phenazine methosulfate (Sigma), 20 mM succinate, and 1 mg/ml nitrotetrazolium blue chloride; Complex IV (cytochrome c oxidase): 50 mM potassium phosphate pH 7.4, 1 mg/ml 3′,3′-diaminobenzidine tetrahydrochloride hydrate (Sigma), 24 units/ml catalase from bovine liver (Sigma), 1 mg/ml cytochrome c from equine heart (Sigma), and 75 mg/ml sucrose [82].

Complexome profiling

Mitochondria from D. melanogaster were analyzed by complexome profiling [83]. Isolated mitochondria (0.2 mg) were solubilized with 6 g digitonin/g protein in 50 mM NaCl, 5 mM 6 aminohexanoic acid, 1 mM EDTA, 50 mM imidazole/HCl, pH 7.0. After centrifugation at 22,000 X g for 20 min at 4°C, the supernatant was supplemented with Coomassie brilliant blue G250 and proteins were separated by 4-16% gradient BN-PAGE. Digitonin solubilized mitochondrial proteins from bovine heart were loaded as molecular mass standards. Gel lanes were cut into 60 slices, transferred to a 96-well filter microtiter plate (Millipore), and destained in 50% (v/v) methanol, 50 mM ammonium bicarbonate. After destaining, in-gel digestion with trypsin was performed. Tryptic peptides were separated by liquid chromatography and analyzed by tandem mass spectrometry (LC-MS/MS) in a Q-Exactive 2.0 Orbitrap Mass Spectrometer (2.8 SP1) equipped with an Easy nLC1000 nano-flow ultra-high-pressure liquid chromatography system (Thermo Fisher Scientific) at the front end. Thermo Scientific Xcalibur 3.1 Software Package was used for data recording. MS RAW data files were analyzed using MaxQuant (version 1.5.0.25). The extracted spectra were matched against the Drosophila melanogaster Uniprot Reference Sequence database (release 2020_04). Database searches were done with 20 ppm match tolerances. Trypsin was selected as the protease with two missed cleavages allowed. Dynamic modifications included N-terminal acetylation and oxidation of methionine. Cysteine carbamidomethylation was set as a fixed modification. Keratins, and trypsin were removed from the list. The abundance of each protein was determined by label-free quantification using the composite intensity based absolute quantification (iBAQ) values determined by MaxQuant analysis and was corrected for loading and MS sensitivity variations between samples based on the total iBAQ value for all detected complex V subunits. Gel migration profiles were created for each protein and normalized to the maximum abundance. Profiles of the identified mitochondrial proteins were hierarchically clustered by distance measures based on Pearson correlation coefficient (uncentered) and the average linkage method using the NOVA software package v0.5 [84]. The visualization and analysis of the heatmaps representing the normalized abundance in each gel slice by a three-color code gradient (black/yellow/red) were done using Microsoft Excel 2019 and Graph Pad Prism 8.4.3. The mass calibration for the BN-PAGE gels was performed as previously described [85]. Membrane proteins were calibrated using the well-known molecular masses of respiratory chain complexes and supercomplexes from bovine heart mitochondria. The soluble proteins were, however, calibrated using the following set of Drosophila proteins: ATPB (51 kDa), malate dehydrogenase (72 kDa, dimer), citrate synthase (100 kDa, dimer), ETFA/B (122, heterodimer), heat shock protein 60 (410 kDa, heptamer), ALDH7A1 (675 kDa, dodecamer).

Western blot and immunodetection

BN-PAGE gels were transferred to PVDF membranes in Dunn carbonate buffer (10 mM NaHCO3, 3 mM Na2CO3) applying a constant current of 300 mA at 4°C for 1 hour using a Mini Trans-Blot® Cell (Bio-Rad). For the immunodetection of specific protein targets, blotted PVDF membranes were blocked in 5% skimmed milk in PBS-T (0.1% Tween-20) at room temperature for 1 hour and then incubated overnight with primary antibodies diluted in 3% BSA in PBS-T overnight at 4°C. PVDF membranes were washed three times with PBS-T for 10 minutes, incubated with the secondary HRP conjugated antibody for 1 hour at room temperature and washed three times with PBS-T for 10 minutes. Chemiluminescent signals were recorded using an Alliance Mini HD9 (UVITEC). Antibodies used are listed in the Key Resource Table. The primary antibodies against D. melanogaster UQCR-C2 and SdhA were a kind gift of Dr. Edward Owusu-Ansah (Columbia University, NY).

High-resolution respirometry

To measure oxygen consumption individuals were homogenized on ice in respiration buffer (120 mM sucrose, 50 mM KCl, 20 mM Tris-HCl, 4 mM KH2PO4, 2 mM MgCl2, 1 mM EGTA, 1% fatty acid-free BSA, pH 7.2). Homogenates were loaded in the chamber of an O2k-HRR (High Resolution Respirometer, Oroboros Instruments) Complex I-linked respiration was measured at saturating concentrations of malate (2 mM), glutamate (10 mM), proline (10 mM) and ADP (2.5 mM). Afterwards, complex II-linked respiration was assessed using 10 mM succinate to the reaction after inhibition of complex I with rotenone (1.25 µM).

Analysis of MRC enzymatic activities

The activities of mitochondrial respiratory chain complexes and citrate synthase (CS) were measured using kinetic spectrophotometric assays as described [81].

Statistical analysis

Statistical analysis was performed with GraphPad Prism Software, version 8.2.1. Statistical tests and significance are described in the figure captions.

Data availability

Complexome profiling data will be deposited in the ComplexomE profiling DAta Resource (CEDAR) repository [86]. Data will be available upon manuscript acceptance.

Supporting information